\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-16 paper.tex
Notes
Updates from Overleaf

The exhastive recording of full inheritance intervals from generation
to generation leads to a rapid build-up of information, much of which
may not be relevant to the genetic ancestry of the current population.
This redundancy can be periodically removed using the ``simplify'' algorithm,
described in detail by \citet{kelleher2018efficient}. Simplification
allows efficient subsetting of ARGs (removing both samples and certain
levels of detail about recombination, as discussed in Section~\ref{sec-ARG-simplification}),
and is also the key enabling factor when incorporating propective ARGs into
forward-time simulation.
The core algorithm involves the concept of
``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}:
the genomic intervals ancestral to a set of samples,
scattered across lineages (the edges of the graph)
at some time in the past. In the next section, we outline how
tracing ancestral material allows the identification and removal of nodes
and inheritance intervals, such that
only genetic ancestry relevant to the current population is retained.

@@

To make the following discussions precise, we provide
a concrete definition and explicit process for local tree extraction.
We can describe a local tree as an
oriented tree~\cite[p.\ 461]{knuth11combinatorial},
a sequence of integers $\pi_1, \pi_2\dots$ where $\pi_u$ denotes the parent of
node $u$~\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
If $u$ is a root
% I'm not sure if this reachability stuff is a helpful clarification, and whether it'll just
% confuse people. It depends on some tricky details.
(or not reachable from the samples)
in the local tree at position $x$, then we adopt the convention that $\pi_u = -1$.
% This sentence needs another pass
A node in an oriented tree can have
zero, one,
or many children.
To recover the local tree $\pi$ at position $x$, we begin by
setting $\pi_u = -1$ for each $u\in N$. Then, for each sample
node in $S$ we trace its path rootwards through the
ARG for position $x$, and record this path in $\pi$.
Specifically, at a given node $u$,
we find an edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
$\pi_c \leftarrow p$. We then set $u \leftarrow p$, and repeat
until either $\pi_u \neq -1$ (indicating we have traversed this section
of the ARG already on the path from another sample) or there
is no matching outbound edge (indicating we are at a root).

@@

This is not the prevailing view, however.
\cite{kuhner2017consensus} argue that the
``interval-tree'' representation of
an ARG (the local trees and the genome intervals they cover)
``does not contain all of the information in the underlying ARG: it lacks the
number of recombinations occuring at each site, the times at which
recombinations occurred, and the specific sequences involved as recombination
partners.''
\cite{shipilina2023origin} discuss the same ideas,
and note that the
``full ARG...~contains more information than the series of tree
sequences along the genome''.
These statements that an ARG
contains more information than its local trees is only true
if we use a representation of the local trees that discards
information about those local trees that can be derived from the ARG.
The internal nodes of evolutionary trees are usually
considered to be \emph{unlabelled}
because they most often correspond
to anonymous hypothetical ancestors.
Internal ARG nodes may likewise correspond to anonymous hypothetical
ancestors, but if we do not label the tree nodes we lose the fact that
the \emph{same} ancestors persist across multiple trees.
Besides having labelled nodes, the local trees must contain
unary nodes marking recombinations
and other processes that lead to lineages passing through a given
ARG node without coalescence in the local trees
(see Section~\ref{sec-locally-unary-edges}).
If the local trees contain labelled internal nodes, they precisely
describing the corresponding path through the graph for each
genome interval and there is exactly as
much information in them as in an ARG.

These properties have interesting consequences when we
consider the inference strategy of estimating local trees
% JK: Relate is inferring trees independently within segments,
% it's just using a dynamic algorithm to decide what those segments
% are, right? That's how it differs from things like Espalier which
% assume a set of input recomb breakpoints.
independently within genome segments,
and subsequently stitching them together into an ARG.
Two recent methods take this approach,
\relate~\citep{speidel2019method} and
\espalier~\citep{rasmussen2022espalier},
and have different strategies for reconciling the local trees into an ARG.
\relate\ first infers local tree topologies, and then identifies
edges that persist from tree to tree by reasoning about the sets of samples
subtended. It does not attempt to be exhaustive.
\espalier, on the other hand, is specifically interested
in the details of recombination events (among a small number of samples)
and therefore attempts to infer
the precise subtree prune-and-regraft (SPR)
operations~\citep{hein1990reconstructing,song2003on,song2006properties}
induced by recombination.
Inferring the SPRs between leaf-labelled trees is
NP-hard~\citep{hein1996complexity,allen2001subtree,bordewich2005computational},
but it is unclear what the complexity is when there
is a degree of internal node sharing between trees.

@@

\section{Levels of simplification}
\label{sec-ARG-simplification}
ARG simplification is a powerful tool, and can do far more
than removing unreachable nodes and edges in a prospective ARG
and resolving ancestral material in an eARG.
In general, we can think of
ARG simplification as the process
of removing nodes and re-writing edges (and their inheritance annotations)
to remove various types of redundancy.
The redundancy that we are interested in
% We are interested in both graph and local unary, so qualifying this
% would just make it less readable, IMO
revolves around the presence of unary nodes (see previous section).
We illustrate this successive removal of redundancy
through a series simplification steps
in Fig.~\ref{fig-simplification}.

@@

The ARG in Fig.~\ref{fig-simplification}A is the output of a
backwards-time Wright-Fisher simulation for a sample of two diploid
individuals (population size $N=10$), and follows a similar process
to the methods described by~\cite{nelson2020accounting}.
As we proceed backwards in time, generation by generation, the
extant lineages chose parents randomly.
With a certain probability recombination occurs, and the ancestral
material of a lineage is split between the two parental
genomes. Local coalesence occurs
when lineages with overlapping ancestral material choose the same
parent genome.
Note that in this simulation we do not explicitly
model recombination \emph{events} via an ARG node, but simply record
the \emph{outcome} of a recombination via outbound edges to
the parent's two genomes.
Thus, a recombinant node may also correspond to a coalescence
(e.g.~node \noderef{g} in Fig.~\ref{fig-simplification}).
The distinction of
using a single node to represent
a recombination event (e.g.~Fig~\ref{fig-ancestry-resolution}),
or two to represent the parent genomes is usually not important
(either is possible in the gARG encoding)
and the most convenient approach will vary by application.
Note also that polytomies are a natural feature of such
a Wright-Fisher model
(e.g.~node \noderef{k} in Fig.~\ref{fig-simplification}).

Fig.~\ref{fig-simplification} shows four ARGs (A--D), with
the graph depicted on the left and the corresponding sequence
of local trees on the right. The graph visualisations have
some novel features, which require some explanation.
Firstly, edge weights (the thickness of the lines joining
nodes) correspond to the length of the inheritance intervals
they are annotated with. This allows us to distinguish
edges that persist across many local trees from those that are
less influential (contrast the edge
$(\noderef{g}, \noderef{h})$
with $(\noderef{g}, \noderef{i})$
in Fig.~\ref{fig-simplification}A).
% FIXME this needs some work
Secondly, node colours denote the number of parents that they
have in the graph, allowing us to easily see roots (those
with zero parents), recombinanants (those with two parents)
and more complex situations arising from simplification (see below).
Thirdly, the shading intensity of a node denotes the
fraction of the node's span (the length of genome in which it
is reachable from the samples in the local trees)
over which it is coalescent (has more than one child).
% This is useful why? I can't come up with a quick way of saying
% this. In some senses it's a measure of identifiability?
% JK: Hmm, not sure we want to jump into inferability here?
All these features will impact what kind of an ARG we can
hope to infer from sampled genomes.

@@

% % Do downstream applications actually make use of such full ARGs?
Besides the inherent limitations that exist on inferring fully
precise ARGs from data,
we should also consider the value that such precise estimates provide
for downstream applications.
To date, few downstream applications using estimated ARGs
make use of complete recombination information.
Most applications work by examining local trees independently.
For example, the \relate\ selection test~\citep{speidel2019method}
obtains $p$-values by computing clade size probabilities conditional on the timing of coalescence events in a given local tree.
In their method
for estimating dispersal rates and the locations of genetic
ancestors,
\citep{osmond2021estimating} downsample trees along the genome
so that they can be regarded as approximately independent.
The SIA method for detecting selection~\citep{hejase2022deep}
encodes local trees as a set of lineage counts at discrete
time intervals, and uses these as feature for a
type of machine learning algorithm
that takes ``temporal'' correlations into account. Thus,
while SIA does use some of the information about local tree correlation,
clearly much of the detail about recombination
events in an ARG is lost.
The \texttt{tsdate} algorithm (and related naive estimator
of ancestral location) uses much more between-tree
information~\citep{wohns2022unified}, explicitly using the gARG
encoding to reason about ancestral nodes.
The main application for fully precise ARGs thus far has been
to compute a likelihood under the coalescent with
recombination~\citep[e.g.][]{kuhner2000maximum,mahmoudi2022bayesian},
which currently requires the details of all recombination
events to be known.

@@

\section{Ancestral graphs: a brief history}
\label{sec-arg-history}
The coalescent~\citep{kingman1982coalescent,kingman1982genealogy,
hudson1983testing, tajima1983evolutionary} models the ancestry of a sample of
genomes under an idealised population model, and provides the theoretical
underpinning for much of contemporary population genetics.
It is a stochastic process, where each random realisation
is a genealogical tree describing the genetic ancestry of the sample.
Numerous extensions to the model have been
proposed~\citep{hudson1990gene,hein2004gene,wakely2008coalescent},
incorporating many evolutionary processes.
\citet{hudson1983properties}
first incorporated recombination into the coalescent process,
providing several fundamental analytical results
and describing the basic simulation algorithm, still in
widespread use~\citep{hudson2002generating,kelleher2016efficient,
baumdicker2021efficient}.
In the 1990s, Griffiths and colleagues revisited the
coalescent with recombination from a different perspective,
formulating it as a stochastic process where each realisation
is encoded as a graph~\citep{griffiths1991two,ethier1990two,
griffiths1996ancestral,griffiths1997ancestral}.
They referred to both the stochastic process and
its random realisations as the Ancestral Recombination Graph (ARG).
Although mathematically equivalent, it is
important to note that the Griffiths and Hudson formulations of
the coalescent with recombination are not identical;
in particular, a direct implementation of the ARG process
as originally described requires exponential time to simulate
(see Appendix~\ref{big-and-little-arg} for details). However, ARGs provided a way
to reason about and infer recombinant ancestry as a single object,
which was not possible within Hudson's framework, which emphasised
the collection of local trees along the genome resulting from recombination.

>>>>

The exhaustive recording of full inheritance intervals from generation
to generation leads to a rapid build-up of information, much of which
may not be relevant to the genetic ancestry of the current population.
This redundancy can be periodically removed using the ``simplify'' algorithm,
described in detail by \citet{kelleher2018efficient}. Simplification
allows efficient subsetting of ARGs (removing both samples and certain
levels of detail about recombination, as discussed in Section~\ref{sec-ARG-simplification}),
and is also the key enabling factor when incorporating propective ARGs into
forward-time simulation.
The core algorithm involves the concept of
``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}:
the genomic intervals ancestral to a set of samples,
scattered across lineages (the edges of the graph)
at some time in the past. In the next section, we outline how
tracing ancestral material allows the identification and removal of nodes
and inheritance intervals, such that
only genetic ancestry relevant to the current population is retained.

@@

% TODO Need to make clear here what came from where. Basic combinatorial object
% cite Knuth. Application to genetic ancestry, sim papers. Need to look
% up what's said in favour of this integer based approach in which paper
% and refer to tat.
To make the following discussions precise, we provide
a concrete definition and explicit process for local tree extraction.
We can describe a local tree as an
oriented tree~\cite[p.\ 461]{knuth11combinatorial},
a sequence of integers $\pi_1, \pi_2\dots$ where $\pi_u$ denotes the parent of
node $u$~\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
If $u$ is a root
% I'm not sure if this reachability stuff is a helpful clarification, and whether it'll just
% confuse people. It depends on some tricky details.
(or not reachable from the samples)
in the local tree at position $x$, then we adopt the convention that $\pi_u = -1$.
% TODO Give example of an oriented tree from one of the figs?
% TODO
% A node in an oriented tree can have
% zero, one,
% or many children.

To recover the local tree $\pi$ at position $x$, we begin by
setting $\pi_u = -1$ for each $u\in N$. Then, for each sample
node in $S$ we trace its path rootwards through the
ARG for position $x$, and record this path in $\pi$.
Specifically, at a given node $u$,
we find an edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
$\pi_c \leftarrow p$. We then set $u \leftarrow p$, and repeat
until either $\pi_u \neq -1$ (indicating we have traversed this section
of the ARG already on the path from another sample) or there
is no matching outbound edge (indicating we are at a root).

@@

This is not the prevailing view, however.
\cite{kuhner2017consensus} argue that the
``interval-tree'' representation of
an ARG (the local trees and the genome intervals they cover)
``does not contain all of the information in the underlying ARG: it lacks the
number of recombinations occuring at each site, the times at which
recombinations occurred, and the specific sequences involved as recombination
partners.''
\cite{shipilina2023origin} discuss the same ideas,
and note that the
``full ARG...~contains more information than the series of tree
sequences along the genome''.
These statements that an ARG
contains more information than its local trees is only true
if we use a representation of the local trees that discards
information about those local trees that can be derived from the ARG.
The internal nodes of evolutionary trees are usually
considered to be \emph{unlabelled}
because they most often correspond
to anonymous hypothetical ancestors.
Internal ARG nodes may likewise correspond to anonymous hypothetical
ancestors, but if we do not label the tree nodes we lose the fact that
the \emph{same} ancestors persist across multiple trees.
Besides having labelled nodes, the local trees must contain
unary nodes marking recombinations
and other processes that lead to lineages passing through a given
ARG node without coalescence in the local trees
(see Section~\ref{sec-locally-unary-edges}).
If the local trees contain labelled internal nodes, they precisely
describe the corresponding path through the graph for each
genome interval and there is exactly as
much information in them as in an ARG.

These properties have interesting consequences when we
consider the inference strategy of estimating local trees
% TODO rephrase this. Ana doesn't like the way it implies that the
% trees are independent, because info from mutations outside the
% window are shared. However, each tree *inference* is independent,
% whether information is used multiple times in different trees or not
independently within genome segments,
and subsequently stitching them together into an ARG.
Two recent methods take this approach,
\relate~\citep{speidel2019method} and
\espalier~\citep{rasmussen2022espalier},
and have different strategies for reconciling the local trees into an ARG.
\relate\ first infers local tree topologies, and then identifies
edges that persist from tree to tree by reasoning about the sets of samples
subtended. It does not attempt to be exhaustive.
\espalier, on the other hand, is specifically interested
in the details of recombination events (among a small number of samples)
and therefore attempts to infer
the precise subtree prune-and-regraft (SPR)
operations~\citep{hein1990reconstructing,song2003on,song2006properties}
induced by recombination.
Inferring the SPRs between leaf-labelled trees is
NP-hard~\citep{hein1996complexity,allen2001subtree,bordewich2005computational},
but it is unclear what the complexity is when there
is a degree of internal node sharing between trees.

@@

\section{Levels of simplification}
\label{sec-ARG-simplification}
ARG simplification is a powerful tool, and can do far more
than removing unreachable nodes and edges in a prospective ARG
and resolving ancestral material in an eARG.
In general, we can think of
ARG simplification as the process
of removing nodes and re-writing edges (and their inheritance annotations)
to remove various types of redundancy.
The redundancy that we are interested in
% We are interested in both graph and local unary, so qualifying this
% would just make it less readable, IMO
revolves around the presence of unary nodes (see previous section).
We illustrate this successive removal of redundancy
through a series of simplification steps
in Fig.~\ref{fig-simplification};
the graph is shown on the left and the corresponding sequence
of local trees on the right.

@@

The ARG in Fig.~\ref{fig-simplification}A is the output of a
backwards-time Wright-Fisher simulation for a sample of two diploid
individuals (population size $N=10$), and follows a similar process
to the methods described by~\cite{nelson2020accounting}.
As we proceed backwards in time, generation by generation, the
extant lineages chose parents randomly.
With a certain probability recombination occurs, and the ancestral
material of a lineage is split between the two parental
genomes. Local coalescence occurs
when lineages with overlapping ancestral material choose the same
parent genome.
Note that in this simulation we do not explicitly
model recombination \emph{events} via an ARG node, but simply record
the \emph{outcome} of a recombination via outbound edges to
the parent's two genomes.
Thus, a recombinant node may also correspond to a coalescence
(e.g.~node \noderef{g} in Fig.~\ref{fig-simplification}).
The distinction of
using a single node to represent
a recombination event (e.g.~Fig~\ref{fig-ancestry-resolution}),
or two to represent the parent genomes is usually not important
(either is possible in the gARG encoding)
and the most convenient approach will vary by application.
Note also that polytomies are a natural feature of such
a Wright-Fisher model
(e.g.~node \noderef{k} in Fig.~\ref{fig-simplification}).

The graph visualisations in Fig.~\ref{fig-simplification} have
some novel features, which require some explanation.
Firstly, edge weights (the thickness of the lines joining
nodes) correspond to the length of the inheritance intervals
they are annotated with. This allows us to distinguish
edges that persist across many local trees from those that are
less influential (contrast the edge
$(\noderef{g}, \noderef{h})$
with $(\noderef{g}, \noderef{i})$
in Fig.~\ref{fig-simplification}A).
% FIXME this needs some work
Secondly, node colours denote the number of parents that they
have in the graph, allowing us to easily see roots (those
with zero parents), recombinants (those with two parents)
and more complex situations arising from simplification (see below).
Thirdly, the shading intensity of a node denotes the
fraction of the node's span (the length of genome in which it
is reachable from the samples in the local trees)
over which it is coalescent (has more than one child).
% This is useful why? I can't come up with a quick way of saying
% this. In some senses it's a measure of identifiability?
% JK: Hmm, not sure we want to jump into inferability here?
All these features will impact what kind of an ARG we can
hope to infer from sampled genomes.

@@

% % Do downstream applications actually make use of such full ARGs?
Besides the inherent limitations that exist on inferring fully
precise ARGs from data,
we should also consider the value that such precise estimates provide
for downstream applications.
To date, few downstream applications using estimated ARGs
make use of complete recombination information.
Most applications work by examining local trees independently.
For example, the \relate\ selection test~\citep{speidel2019method}
obtains $p$-values by computing clade size probabilities conditional on the timing of coalescence events in a given local tree.
In their method
for estimating dispersal rates and the locations of genetic
ancestors,
\cite{osmond2021estimating} downsample trees along the genome
so that they can be regarded as approximately independent.
The SIA method for detecting selection~\citep{hejase2022deep}
encodes local trees as a set of lineage counts at discrete
time intervals, and uses these as feature for a
type of machine learning algorithm
that takes ``temporal'' correlations into account. Thus,
while SIA does use some of the information about local tree correlation,
clearly much of the detail about recombination
events in an ARG is lost.
The \texttt{tsdate} algorithm (and related naive estimator
of ancestral location) uses much more between-tree
information~\citep{wohns2022unified}, explicitly using the gARG
encoding to reason about ancestral nodes.
The main application for fully precise ARGs thus far has been
to compute a likelihood under the coalescent with
recombination~\citep[e.g.][]{kuhner2000maximum,mahmoudi2022bayesian},
which currently requires the details of all recombination
events to be known.

@@

\section{Ancestral graphs: a brief history}
\label{sec-arg-history}
The coalescent~\citep{kingman1982coalescent,kingman1982genealogy,
hudson1983testing, tajima1983evolutionary} models the ancestry of a sample of
genomes under an idealised population model, and provides the theoretical
underpinning for much of contemporary population genetics.
It is a stochastic process, where each random realisation
is a genealogical tree describing the genetic ancestry of the sample.
Numerous extensions to the model have been
proposed~\citep{hudson1990gene,hein2004gene,wakely2008coalescent},
incorporating many evolutionary processes.
\citet{hudson1983properties}
first incorporated recombination into the coalescent process,
providing several fundamental analytical results
and describing the basic simulation algorithm, still in
widespread use~\citep{hudson2002generating,kelleher2016efficient,
baumdicker2021efficient}.
In the 1990s, Griffiths and colleagues revisited the
coalescent with recombination from a different perspective,
formulating it as a stochastic process where each realisation
is encoded as a graph~\citep{griffiths1991two,ethier1990two,
griffiths1996ancestral,griffiths1997ancestral}.
They referred to both the stochastic process and
its random realisations as the Ancestral Recombination Graph (ARG).
Although mathematically equivalent, it is
important to note that the Griffiths and Hudson formulations of
the coalescent with recombination are not identical;
in particular, a direct implementation of the ARG process
as originally described requires exponential time to simulate
(see Appendix~\ref{sec-big-and-little-arg} for details). However, ARGs provided a way
to reason about and infer recombinant ancestry as a single object,
which was not possible within Hudson's framework, which emphasised
the collection of local trees along the genome resulting from recombination.
